\title{\bf Distance Ladder}

\section{Basics \& Nomenclature}

Determination of distance is a fundamental problem in astronomy, to
which is owed numerous breakthroughs. The distance to nearby galaxies
has been established through a calibrated {\it distance ladder},
refined over the past century.

The distance ladder begins with the measurement of {\it parallax},
which is the apparent motion of sufficiently nearby stars due to the
motion of the Earth around the Sun. Precise measurements of stars over
time in a fixed reference frame yield {\it proper motions} and
parallax-based distances. Most recently, the Gaia satellite has
provided a large sample of stars with measured parallaxes reaching
across our Galaxy.

To reach distances outside our galaxy, we must rely on less direct
measurements of distance. Usually this involves finding objects or
phenomena that have pairs of observables with a known relation but
with different dependences on distance.

Many of these techniques fall into the {\it standard candle} category:
a luminosity is estimated by some means, and a flux is measured, and
the distance is inferred from the inverse-square law $f=L/4\pi D^2$.
The most prominent standard candles are:
\begin{itemize}
\item {\it RR Lyrae} variable stars. These stars are relatively low mass
  horizontal branch stars. They vary due to the $\kappa$ mechanism,
  which is an instability associated with the reaction of the
  atmospheric opacity to changes in pressure and density that causes
  oscillations in stellar size and luminosity.  They have periods of
  order a day. They are prominent in globular clusters but can also be
  found throughout the Milky Way. Historically they have been most
  useful for determining distances within the Galaxy. However, the
  advent of Gaia and the James Webb Space Telescope opens up the
  possibility of observing these stars in nearby galaxies.
\item {\it Cepheid} variable stars. These stars are high mass and
  luminous.  They also vary due to the $\kappa$ mechanism but have
  longer periods. Historically, they have provided one of the most
  direct links between local distance scales and distant galaxies.
\item {\it Tip of the Red Giant Branch} stars. The red giant branch is
  commences when core hydrogen burning ends; at this point, shell
  hydrogen burning commences and the star becomes rapidly brighter and
  much larger. At a certain point the helium ash at the core ignites,
  moving the star to its next phase, the smaller and less luminous
  horizontal branch. The core ignites rapidly at a specific stage in
  evolution, so that at a given metallicity red giant stars all end
  this stage at the same luminosity. This leads to a rapid fall-off in
  the luminosity distribution of RGB stars and a clear tip of the red
  giant branch that can be used as a standard candle.
\item {\it Type Ia supernovae}. These supernovae result from the
  nuclear detonation of white dwarfs, probably due to accretion from
  or collision with a binary companion. The Phillips relation connects
  the time scale of the supernova light curve to the luminosity of the
  supernova. These sources were used to provide the first definitive
  evidence for the existence of cosmic acceleration.
\end{itemize}
Historically, a number of other distance measurements have also been
used that could be categorized as standard candle distances. Some
methods have relied on other supernova types. Other methods have
generally capitalized on scaling relations of galaxies, such as the
Tully-Fisher relation for spirals relating luminosity and circular
velocity, and the Fundamental Plane for ellipticals relating
luminosity, size, and velocity dispersion. The difficulty in
implementing these methods in a way that is free of systematic errors
generally has led to the use of supernovae for the most recent
measurements of the distance ladder.

Several standard candle techniques prove useful, or may soon do so. I
will discuss eclipsing binaries, masers, and gravitational waves:
\begin{itemize}
\item Some binary stars, like Algol, eclipse each other during their
orbit. Their light curve alone yields a period $P$ and their radii
relative to the orbit size, $r_1/a$ and $r_2/a$ for any given
eccentricity and phase of eclipse. If both their spectra can also be
measured, this can yield velocities, which yields enough information
to determine their total mass $M$, eccentricity $e$, and the
line-of-sight semimajor axis of the orbit $a \sin i$. The mass and
semimajor axis are related by:
\begin{equation}
M = \frac{4\pi^2 a^3}{GP^2}.
\end{equation}
For eclipsing binaries $i\approx 90^\circ$.  Thus, from gravitational
and geometric considerations, you can determine the sizes of the
stars. Spectra can yield constraints on their surface
temperatures. The sizes and temperatures determines the luminosities,
which combined with fluxes determines the distance.
\item Under certain conditions the supermassive
black holes at the centers of galaxies may contain H$_2$O
masers. Under even more special conditions, these systems can be
sufficiently clean that one may measure velocities and accelerations
over time of clouds in circular orbits, which yields a physical
distance. Combined , this combination yields
the mass of the black hole and the distance. Only one clean enough
system is known, about 8 Mpc away in NGC 4258.
\item Colliding compact objects whose gravitational waves are detected can
yield constraints on distance from the gravitational waveform, because
the total mass of the system is related to both the waveform frequency
and the total energy. If the redshift is known from an electromagnetic
counterpart, then the rest-frame waveform frequency can be determined,
and the object can be placed on the redshift-distance relation.
\end{itemize}

A major application of these methods is to compare the redshifts and
distances of local galaxies. This comparison led Hubble to the Hubble
Law:
\begin{equation}
v = H_0 d
\end{equation}
where $v$ is velocity, $H_0$ is the Hubble constant, and $d$ is the
distance. Gravitational attraction causes galaxies to move toward each
other with respect to this flow, with {\it peculiar velocities} of
100s of km s$^{-1}$. For example, M31 is moving toward us at 400 km
s$^{-1}$, and large clusters, such as Virgo and Coma, have internal
motions and thus peculiar velocities of 1000s of km s$^{-1}$. Thus,
determinations of the Hubble constant need to use galaxies
sufficiently far to reduce this effect. Alternatively, they need to
estimate and correct for these peculiar velocities, using the density
field of galaxies itself to estimate the magnitudes and directions of
the expected velocities.

The Hubble Law is a fundamental measurement of the nature of the
universe we are in, but it also is a tool to map the universe on
larger scales, since measuring Doppler shifts is relatively easy
compared to galactic distances.

\section{Commentary}

Briefly, the most precise current estimates of the local Hubble
constant come from the \citet{riess16a} group. These use a number of
Milky Way Cepheids with parallax distances, along with eclipsing
binaries and NGC 4258, to anchor extragalactic Cepheids in galaxies
with SNe Ia, and use those to anchor a much larger SN Ia data set. As
of 2018, the results for $H_0$ from these analyses tend to be
considerably higher (several sigma) than those inferred from cosmic
microwave background and large scale structure measurements, using
standard cosmological models to extrapolate to $z=0$. The cause of
this discrepancy is not known.

The current ladder to a precise $H_0$ is strongly dependent on the
Cepheid step. There are several important issues with the
interpretation of the Cepheid distance scale, the most prominent of
which is that the period-luminosity relation is metallicity
dependent. The LMC has a much lower metallicity than the other
galaxies in the analysis, and so if improperly modelled the
metallicity dependence can cause a bias.

\citet{beaton16a} present an alternative path using different
indicators (RR Lyrae and TRGB) to calibrate the SN Ia, both of which
tend to trace an older, more metal poor population even within metal
rich galaxies. This path may soon become competitive. This second
indepedent measurement will be critical in assessing the existing
discrepancy.

The detection of much more distant masing systems suitable for an
analysis similar to that of NGC 4258, or a sufficient number of
gravitational wave sources, could provide a ``one step'' alternative
to the distance ladder for the purpose of determining the cosmological
constant.

\section{Key References}

\begin{itemize}
  \item
    {\it A 2.4\% Determination of the Local Value of the Hubble
    Constant}, \citet{riess16a}
  \item
    {\it The Carnegie-Chicago Hubble Program. I. An Independent
    Approach to the Extragalactic Distance Scale Using Only Population
    II Distance Indicators}, \citet{beaton16a} 
\end{itemize}

\section{Order-of-magnitude Exercises}

\begin{enumerate} 
\item Estimate age of the universe from the Hubble constant assuming
    constant expansion.
\item What is the maximum parallax at 1 pc, 10 pc, 1 kpc, and 1 Mpc?
    Assuming you can get good positions to with a few hundredths of a
    FWHM, what size telescopes do you need so that you can measure
    parallax at these distances?
\item If the typical peculiar velocity of a galaxy is $\sim 300$ km
    s$^{-1}$, at what redshift does its contribution to the distance
   error drop below 5\%?
\item Based on their luminosities, what are the relative distances to
   which you should be able to determine distances to RR Lyrae,
   TRGB stars, Cepheids, and SN Type Ia?
\end{enumerate} 

\section{Numerics and Data Exercises}

\begin{enumerate}
\item \citet{clementini18a}
\end{enumerate}

\bibliographystyle{apj}
\bibliography{exex}  
