\documentclass[11pt, preprint, maxnames=9]{aastex}
\usepackage{hyperref}
\usepackage{rotating}
 
\setlength{\footnotesep}{9.6pt}

\newcounter{thefigs}
\newcommand{\fignum}{\arabic{thefigs}}

\newcounter{thetabs}
\newcommand{\tabnum}{\arabic{thetabs}}

\newcounter{address}

\begin{document}

\title{\bf Light I}

\section{Basics \& Nomenclature}

In astrophysical parlance, the {\it specific intensity} $I_\nu$ is the
flux of electromagnetic energy across a surface from a particular
direction, per unit area, per unit solid angle, per unit time, per
unit frequency. The specific intensity can also be expressed per unit
wavelength, and denoted in this case $I_\lambda$. For example, units
may be erg s$^{-1}$ cm$^{-2}$ Hz$^{-1}$ arcsec$^{-2}$ or erg s$^{-1}$
cm$^{-2}$ \AA$^{-1}$ arcsec$^{-2}$.  The term {\it surface brightness}
is often also used to denote the specific intensity.

The quantities $I_\nu$ and $I_\lambda$ are related:
\begin{eqnarray}
I_\nu(\nu) |{\rm d}\nu| &=& I_\lambda(\lambda = c/\nu) |{\rm
  d}\lambda| \cr
I_\nu(\nu) &=& \left|\frac{{\rm d}\lambda}{{\rm d}\nu}\right|
I_\lambda = \frac{c}{\nu^2} I_\lambda(\lambda = c/\nu) \cr
I_\lambda(\lambda) &=& \left|\frac{{\rm d}\nu}{{\rm d}\lambda}\right|
I_\nu = \frac{c}{\lambda^2} I_\nu(\nu = c/\lambda).
\end{eqnarray}

A {\it flux density} $f_\nu$ or $f_\lambda$ is the integral of the
specific intensity through the surface integrated over a range of
directions:
\begin{equation}
f_\nu = \int {\rm d}\Omega I_\nu \cos\theta. 
\end{equation}
A special unit for flux density (mostly in use in radio astronomy) is
the Jansky, which is $10^{-23}$ erg s$^{-1}$ cm$^{-2}$
Hz$^{-1}$. Sometimes the flux density is also called the {\it spectral
  energy distribution}.

The electromagnetic wave can also be expressed in terms of
photons. The energy of photons correspond to a wavelength of light:
\begin{equation}
E_\nu = h\nu
\end{equation}

If an emitting object is moving along the line-of-sight to the
observer, the photon's wavelengths are shifted by a factor $(1+z)$
where $z$ is the redshift:
\begin{eqnarray}
\lambda_{\rm obs}  &=& (1+z) \lambda_{\rm emit} \cr
\nu_{\rm obs}  &=& \frac{\nu_{\rm emit}}{1+z}
\end{eqnarray}
The specific intensity and the flux is also altered by this
motion. Under both special and general relativity, the photon density
in phase space is conserved. From this it can be calculated that:
\begin{equation}
  \label{eq:sb_dimming}
  I_{\lambda, {\rm obs}} = 
  \frac{I_{\lambda, {\rm emit}}}{(1+z)^4}
\end{equation}
At small velocities, $z \approx v/c$, but at higher velocities
relativistic corrections are important.

A {\it luminosity} is defined as the total energy emitted from some
source per unit time per unit frequency (or wavelength). It has
typical units of erg s$^{-1}$ Hz$^{-1}$ or erg s$^{-1}$
\AA$^{-1}$. For isotropically emitting sources, the total luminosity
is related to the observed flux as:
\begin{equation}
\label{eq:luminosity_distance}
f_\nu = \frac{L_\nu}{4\pi D^2}
\end{equation}
where $D$ is the distance to the source.

In a cosmological context (relevant at 100s of Mpc distance or so
depending on the required precision), there is not a single
well-defined distance to a source. The directly observable quantity is
the redshift $z$, associated with the Universe's expansion. In this
case $D$ must be the ``luminosity distance,'' defined to satisfy
Equation \ref{eq:luminosity_distance}; its relation to redshift
depends on the cosmological parameters (see \citealt{hogg99cosm}). The
flux dependence on distance in standard, general relativity-based
cosmology is a combination of the dependence of the angular size on
distance times the surface brightness dimming effect.

The flux density or specific intensity can only be measured at some
finite resolution in wavelength or frequency. In optical astronomical
parlance, a common expression of resolution is:
\begin{equation}
R = \frac{\lambda}{\Delta \lambda}
\end{equation}
where $\Delta \lambda$ is the full-width half maximum (FWHM) of the
line spread function (the Green's function response of the system to a
point source).

Optical and infrared instruments that measure the wavelength
dependence of the flux density through dispersal of the light (with
diffraction gratings or prisms in the optical) are typically known as
spectrographs. Existing instruments range in resolution from $R\sim
20$ to $R>100,000$.  At higher energies, in addition to dispersal
techniques, X-ray detectors and position sensitive proportional
counters also often have intrinsic energy sensitivity.  Radio
receivers separate signals into frequencies electronically.

Imaging instruments with wavelength-dependent sensitivity (either due
to the detector or through a band pass filter) can also measure the
wavelength dependence of the flux density or specific
intensity. Typically, this is performed more coarsely $R\sim 5$ though
narrow band systems reach $R\sim 50$ (typically though not quite
always requiring one exposure per filter).

In the optical and infrared the use of such coarse band pass filters
means that the interpretation of the measurements at high precision
requiring knowing the band pass very well. A full interpretation of
the measurement involves also a model of the flux density. The
observations are usually interpreted as the ratio between the signal
received from the object to the signal that would have been received
by some standard source.

This quantity is termed a {\it maggie} and depends on the band
pass. It can be expressed in terms of a model for $f_\nu$ or
$f_\lambda$ of the object and $g_\nu$ or $g_\lambda$ of the standard
source:
\begin{eqnarray}
  \mu_b &=&
  \frac{\int {\rm d}\nu (f_\nu(\nu)/\nu) R_b(\nu)}
       {\int {\rm d}\nu (g_\nu(\nu)/\nu) R_b(\nu)} \cr
       &=&
  \frac{\int {\rm d}\lambda f_\lambda(\lambda) \lambda R_b(\lambda) }
       {\int {\rm d}\lambda g_\lambda(\lambda) \lambda R_b(\lambda)} 
\end{eqnarray}
where $R(\lambda)$ or $R(\nu)$ is the contribution of a photon
entering Earth's atmosphere (or the aperture of a space telescope) to
the output signal of the instrument in band $b$. This formula is the
same for photon-counting and energy-counting devices, though note that
for the latter instrumentalists typically report a different
definition of $R(\lambda)$ or $R(\nu)$, that therefore needs to be
transformed into the aforementioned form.

Typical choices of $g_\nu$ are the spectrum of Vega (which is only
known to a few percent accuracy in the optical and near-infrared) and
the AB system of $g_\nu = 3631$ Jy (the flux density of Vega near 5500
\AA).  All absolute measurements are based on standards whose flux
densities are thought to be known, but only rarely by comparing
directly to Vega (which is too bright) and never to an AB source (they
don't exist). Relative measurements can be calibrated fairly precisely
independently of absolute calibration, though for broad band passes
the colors of the sources need to be accounted for.

The classical astronomical {\it magnitude} is defined as:
\begin{equation}
m = -2.5 \log_{10} \mu
\end{equation}
The bright end of this system corresponds roughly to the original
magnitude system developed by astronomers in the ancient world.

The luminosity is often expressed as the {\it absolute magnitude},
which is defined as the magnitude the object would have were it at
rest with respect to us, at 10 pc distance. For non-cosmologically
distant objects (within a few 10s of Mpc) this can be expressed as:
\begin{equation}
M = m - 5 \log_{10} \left(\frac{D}{10 {\rm ~pc}}\right) = m - {\rm DM}
\end{equation}
where DM is defined as the distance modulus. 

At cosmological distances, two effects are important. First, $D$ must
be the luminosity distance. Second, the observed band pass corresponds
to a different part of the rest frame spectrum of the object for each
different redshift $z$. This effect is known as the $K$-correction,
defined as:
\begin{equation}
M = m - {\rm DM} - K(z; f_\lambda)
\end{equation}
Basically the $K$-correction accounts for the ratio of the observed to
rest frame flux given the shape of the flux density
$f_\lambda(\lambda)$ (it does not depend on the amplitude). It is
common at low redshifts to $K$-correct from the observed bandpass to
the same band pass in the rest frame. However, at higher redshift it
can be more stable to $K$-correct to a bandpass with a closer
effective wavelength to that observed. \citet{hogg02c} describes the
mathematical definition of the $K$-correction in all such cases. 

\section{Commentary}

While the magnitude system was originally used for convenience in
calibration over a large dynamic range in brightness, it retains its
usefulness because of its encapsulation of the response function
$R(\lambda)$.  Sometimes measurements of $\mu$ are expressed in terms
of $f_{\lambda, {\rm eff}}$ at some choice of $\lambda_{\rm eff}$
(e.g. by just multiplying an AB maggie by 3631 Jy and using some
average effective wavelength of the filter), but for broad band passes
either $f_{\lambda, {\rm eff}}$ or $\lambda_{\rm eff}$ is a strong
function of the actual, often unknown $f_\lambda(\lambda)$. $\mu$ and
$m$ do not suffer from this dependence (though interpreting them
physically still can only be done by accounting for the filter curve
and a model for $f_\lambda(\lambda)$).

$K$-corrections depend on knowing the flux density itself, or at least
its wavelength dependence. To estimate that wavelength dependence
requires having a measurement or model. Thus, the $K$-correction is
naturally an uncertain quantity.

\section{Key References}

\begin{itemize}
  \item
    \href{http://adsabs.harvard.edu/abs/2000asqu.book.....C}{
    {\it Allen's Astrophysical Quantities},
      \citet{cox00a}}, Chapter 5
  \item
    \href{http://adsabs.harvard.edu/abs/1999astro.ph..5116H}{
    {\it Distance measures in Cosmology},
      \citet{hogg99cosm}}
  \item
    \href{http://adsabs.harvard.edu/abs/2002astro.ph.10394H}{
    {\it The $K$-correction},
      \citet{hogg02c}}
\end{itemize}

\section{Order-of-magnitude Exercises}

\begin{enumerate}
\item If Vega is about 8 parsecs away, how far away would it be
  visible through the deepest images from the Hubble Space Telescope
  ($m\sim 30$)?

\item Estimate the number of photons per second that enter your eye in
  visible light (4000--7000 \AA) from a star with magnitude $\sim 6$
  (about the faintest visible at a dark site). Assume a nighttime
  pupil diameter of 5 mm.

\item How much does surface brightness dimming change the magnitudes
  per square arcsecond for a galaxy at redshift $z\sim 1$?
\end{enumerate}

\section{Analytic Exercises}

\begin{enumerate}
\item For a Gaussian line spread function of width $\sigma$, what is
  the FWHM?
\item Prove Equation \ref{eq:sb_dimming}, based on the fact that
  photon density in phase space is conserved.
\item Based on Equation \ref{eq:sb_dimming}, how is the angular
  diameter distance related to the luminosity distance? 
\end{enumerate}

\section{Numerics and Data Exercises}

\begin{enumerate}
\item Retrieve a spectrum of a star, a quasar, and a galaxy from the
  Sloan Digital Sky Survey, and plot each of them. 


\bibliographystyle{apj}
\bibliography{exex}  

\end{document}

