\documentclass[11pt, preprint]{aastex}
\usepackage{hyperref}
\usepackage{rotating}
 
\setlength{\footnotesep}{9.6pt}

\newcounter{thefigs}
\newcommand{\fignum}{\arabic{thefigs}}

\newcounter{thetabs}
\newcommand{\tabnum}{\arabic{thetabs}}

\newcounter{address}

% TODO
%
% Links in intro
% Location of class
% Update distance ladder commentary

% Assignments not done:
%  atmosphere OOM and analytic
%  detector OOM #1
%  All analytic in "images"
%  Cosmology OOM and analytic
%  Structure OOM 
%  Structuer analytic

\begin{document}

\title{\bf Extragalactic Astrophysics / PHYS-GA 2051 / Fall 2024 / Syllabus }

~

\noindent This course teaches the astrophysics of galaxies and quasars
at the graduate level.

\noindent You can find the course notes at the
\href{http://blanton144.github.io/exex}{course web site}. Please read
the introduction posted on the web site.

\noindent Useful textbooks are {\it Galaxy Evolution} by Cimatti,
Fraternali, \& Nipoti, and {\it Extragalactic Astronomy and
  Cosmology}, by Peter Schneider. A good fraction of my notes are
drawn from those books.

\noindent Class meets Monday and Wednesday at 11:00am in Room {\bf
  XYZ} of 726 Broadway.

\noindent The classes will proceed as shown on the next page (subject
to revision!).

\noindent The following are the expectations in the course and
classes:

\begin{itemize}

\item {\it Reading}: I expect you to read the provided notes {\it
  before} each class. I am likely to call on you in class to ask 
  specific questions about things I think are especially important.

\item {\it Homework}: Each week I will assign one of the questions in
  the notes we covered, due each week starting Tuesday, Sept 16. I
  encourage you to discuss and work on this together. If any subset of
  you would like to submit as a group, please ask my permission
  beforehand.

\item {\it Homework Presentation}: On Thursday after each exercise is
  due, I will ask one student to describe their solution to the class;
  you won't have warning but I am not expecting a polished
  presentation.

\item {\it Review Paper \& Presentation}: Before Tuesday, Sept 16, I
  will assign you each a topic covering a recent finding in
  extragalactic astrophysics. You will prepare a short review paper
  and a presentation for the class. The paper should be formatted in
  \LaTeX and be about 8 pages of text and figures plus references and
  (if appropriate) figures. Mid-semester a {\it full draft} of this
  paper will be due---meaning, a version you think is done (spoiler
  alert: I will have comments and questions and requests to more fully
  probe issues so it won't be done). I expect to give substantial
  feedback on the draft in preparation for the final version due at
  the semester's end. You will each prepare a 10 minute presentation
  summarizing your review paper.

\end{itemize}

\baselineskip 0pt
\begin{table}
\footnotesize
\begin{tabular}{|c||c|c|}
\hline
{\it Sep.~4} & Inventory & \cr
{\it Sep.~9} & Light I \& II & \cr
{\it Sep.~11} & Telescopes \& Atmosphere & \cr
{\it Sep.~16} & Detectors, Images, Spectra & {\bf Exercise \#1 due} \cr
{\it Sep.~18} & Distance Ladder & \cr
{\it Sep.~23} & Cosmology \& Structure & {\bf Exercise \#2 due} \cr
{\it Sep.~25} & Stellar Populations I & \cr
{\it Sep.~30} & Stellar Populations II & {\bf Exercise \#3 due} \cr
{\it Oct.~2} & Stellar Populations III & \cr
{\it Oct.~7} & Galaxy Populations & {\bf Exercise \#4 due}\cr
{\it Oct.~9} & Stellar Dynamics I & \cr
{\it Oct.~14} & Stellar Dynamics II &  {\bf Exercise \#5 due} \cr
{\it Oct.~16} & Stellar Dynamics III & \cr
{\it Oct.~21} & Stellar Dynamics IV & {\bf Exercise \#6 due} \cr
{\it Oct.~23} & Interstellar Medium I &  \cr
{\it Oct.~28} & Interstellar Medium II  & {\bf Full paper draft due} \cr
{\it Oct.~30} & Dust & \cr
{\it Nov.~4} & Gravitational Lensing I & {\bf Exercise \#7 due} \cr
{\it Nov.~6} & Gravitational Lensing II & \cr
{\it Nov.~11} & Gravitational Lensing III &   {\bf Exercise \#8 due} \cr
{\it Nov.~13} & Groups \& Clusters &   \cr
{\it Nov.~18} & Active Galactic Nuclei I &  {\bf Exercise \#9 due} \cr
{\it Nov.~20} & Active Galactic Nuclei II & \cr
{\it Nov.~25} & Lyman-$\alpha$ Forest & {\bf Exercise \#10 due} \cr
{\it Dec.~2} & Theory of Galaxy Formation I & \cr
{\it Dec.~4} & Theory of Galaxy Formation II & {\bf Exercise \#11 due} \cr
{\it Dec.~8} & Theory of Galaxy Formation III & \cr
{\it Dec.~11} & Future of Extragalactic Astronomy & \cr
{\it Dec.~19} & --- & {\bf Final paper due} \cr
\hline
\end{tabular}
\end{table}

% Other topics
%  - ly-alpha

\end{document}

